Gradual Soundness: Lessons from Static Python

### 2022-01-08 pj-7.1/paper.tex
pj: sec 2, hide ChkList figure, no strong ideas yet for filling it

Parameterized types (generics) are more difficult to fully enforce without
changing the language.
Consider a list datatype \code{List[T]} that is parameterized by the
e type of its elements.
If a function expects lists of integers (\code{List[int]}), then a simple check
for lists is not strong enough to decide whether an incoming value matches the
type~(\cref{f:list-boundary}).
A deeper check is needed to validate the list elements.
But, the default \code{List} type cannot support such a check because untyped
code is free to create lists.

@@

\begin{figure}[t]
  How to enforce type \code{List[int]} simply?
  \begin{itemize}
    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
    \item Show venn diagram of the options?
  \end{itemize}
  \caption{FILL illustrate list boundary and checks}
  \label{f:list-boundary}
\end{figure}

@@

%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least
%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)

@@

As of December 2021, the \SP{} team has converted 541 modules.
Most of these came from the tool (417); the rest
are from hand conversions (124).
These modules are well-mixed with untyped modules in the codebase.
All together, over thirty thousand identifiers cross a type boundary
at compile-time by either entering or leaving a \SP{} module.
Two-thirds of these crossings are exports from \SP{} modules
to untyped modules;
in other words, the typed identifiers are widely used throughout the Instagram web server.
%% data in src/static-python-imports-analysis

>>>>

%% TODO 2022-01-07 List and ChkList are incompatible, need to anticipate somehow ... Chk does not allow any lists!
Parameterized types (generics) are more difficult to fully enforce without
changing the language.
Consider a list datatype \code{List[T]} that is parameterized by the
e type of its elements.
If a function expects lists of integers (\code{List[int]}), then a simple check
for lists is not strong enough to decide whether an incoming value matches the
type.
A deeper check is needed to validate the list elements.
But, the default \code{List} type cannot support such a check because untyped
code is free to create lists.

@@

%\begin{figure}[t]
%  How to enforce type \code{List[int]} simply?
%  \begin{itemize}
%    \item Checking for \code{List} is simple, but loose. It accepts \code{List[str]} and heterogeneous lists etc.
%    \item Can't check for \code{List[int]} by itself, the datatype does not support that.
%    \item Checking for \code{ChkList[int]} is simple, but needs buy-in from users.
%  \end{itemize}
%  2022-01-07: might be an interesting figure idea, showing the two options for enforcing a data structure
%    Q How to enforce data structure without traversing
%     \=> option 1: Shallow
%     \=> option 2: Deep
%  \caption{FILL illustrate list boundary and checks}
%  \label{f:list-boundary}
%\end{figure}

@@

%% 2022-01-06 TODO subtyping for classes C ... matters for casts at least ... just add a Gamma, follow chain up one step at a time until refl.
%% 2022-01-04 KC will look into pbt for nom properties (I don't think there's much to see there!)

@@

As of December 2021, the \SP{} team has converted 541 modules.
Most of these came from the tool (417); the rest
are from hand conversions (124).
These modules are well-mixed with untyped modules in the codebase.
All together, over thirty thousand identifiers cross a type boundary
at compile-time by either entering or leaving a \SP{} module.
Two-thirds of these crossings are exports from \SP{} modules
to untyped modules;
in other words, the typed identifiers are widely used throughout the Instagram web server.
%% data in src/static-python-imports-analysis
%% 2022-01-06 TODO gen-code is configuration system / access control ... what are stats omitting these?!
